Human capital mediates natural selection in contemporary humans

### 2022-03-15 response-to-reviewers-BE.Rmd

---
output: 
  pdf_document:
    toc: false
    latex_engine: xelatex
header-includes:
  - \usepackage{subfig}
  - \hypersetup{colorlinks = true, linkcolor = {blue}, linkbordercolor = {white}}
  - \usepackage{tcolorbox}
  - \newtcolorbox{myquote}{colback=green!5!white, colframe=green!75!black}
  - \renewenvironment{quote}{\begin{myquote}}{\end{myquote}}
mainfont: Baskerville
fontsize: 12pt
---

```{r setup, echo=FALSE}

@@

  if (is.null(conf_int)) conf_int <- "conf.low" %in% names(dfr)

  n_regs <- as.double(n_regs)

  confint_geom_segment <- if (conf_int) confint_geom_segment() else NULL

  dfr %>% mutate(
    !! score_col := pretty_names(!! score_col),
    !! score_col := fct_reorder(!! score_col, estimate, order_abs(order_idx))
  ) %>% 
    rename(p = p.value) %>% 
    ggplot(aes(estimate, {{score_col}}, yend = {{score_col}}, 
               shape = {{fill_col}}, ...)) +
    my_vline + 
    confint_geom_segment +
    geom_point(size = 1.8, aes(color = {{fill_col}})) +
    geom_point(size = 0.7, color = "white", aes(alpha = p < 0.05/33)) +
    scale_color_manual(
                         values = rev(c("lightblue3", "dodgerblue2", "dodgerblue4"))
                       ) +
    scale_shape_manual(values = 1:n_cats + 14) +
    my_labs() +
    my_theme() +
    scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
    scale_alpha_discrete(range = c(1, 0),
                         guide = guide_legend(
                           override.aes = list(
                             shape = c(1, 16),
                             color = "black",
                             alpha = 1,
                             size = 1.8
                           )
                         ))
}

@@

We accept point (6) in particular. On the other hand, we do state that this is a "back of the envelope calculation". We've had pushback from people saying "this is no big deal", so we want to show that at least under some assumptions, yes it might be. Note also that as we say:

"If unmeasured PSEA correlates positively with $\varepsilon$, then the reliability coefficient $\lambda$ will be lower than calculated. However, the proportional effect of natural selection on the PSEA-income relationship will remain the same, so long as $\lambda$ does not differ between the two sets of weighted regressions."

The within-family estimate should remove the worries about exogeneity of p\*, but they are estimated with a smaller sample. We agree that the estimate is high. The bootstrapped 95% confidence interval is from around 0% to 300%, which is not really informative, though it does suggest that genetic nurture is more likely to bias our results downwards than upwards. We've removed the discussion.

> p\. 40 and following: I confess to not following the derivations completely. The authors seem adept at what they do, and I trust that they have made no serious errors.
>
> (1) This is my first encounter with the Karush-Kun-Tucker conditions or the Inada condition.

@@

We have tried to work on this. Unconnected results and minor points have been pushed to footnotes and/or the appendix.

We don't want to go too far, by pretending that the theory came to us fully-formed in our ivory tower, and then we tested it. In fact, the motivating results led us to develop the theory, first informally and then more precisely; only the results in section 4 should be considered a clean test.

> In my opinion, the paper's most important finding is that human capital mediates natural selection. I suspect the authors agree with me on this point, since the title of their manuscript says just that. Still, when reading the paper, one does not get to the topic of human capital and natural selection until p. 9 (and the paper has only 15 pages)! The lengthy discussion of the stratified analysis (Figures 3-5) is interesting, and the authors have edited the text (vs. in the earlier version) to explain that these results are related to predictions of their model, but that discussion is too long and distracts from the main point of the paper. Plus, the stratified analyses are\
> limited and only so much can be learned from them. As the authors acknowledge, "these categories are not necessarily exogenous to the polygenic scores. For example, both in the data and in our theoretical model, age at first live birth is a choice variable, which is endogenous to human-capital-related polygenic." Thus, I recommend that that discussion be shortened and the figures consolidated (the authors should find a way to show the must important results in Figures 3-5 on \~1 page.

We think these analyses are important. First, they motivate our theory, in which education, income, AFLB etc. are indeed endogenous choice variables which are co-determined with fertility. If these mediators were *not* endogenous, we'd need a different theory.

@@

We also think the results matter in their own right. Any theory of contemporary natural selection needs to explain why correlations with fertility are lower at higher levels of income, education, etc. Non-causal explanations would be fine, but theories which don't generate these correlations must be missing something. Note that Fisher, cited by Reviewer 1, used observed correlations quite like ours to differentiate between theories of the fertility-income relationship. Generally, significant correlations have often led to important theoretical literatures, even when the direction of causality is uncertain (e.g. education/wages in labour economics, or democracy/GDP levels in political economy).

As described above, we have shortened Section 2 to focus on the key results of variation by income, education, number of partners, partner presence and AFLB. These empirics now take up just 39 lines of text, plus Figures 3-5. We no longer split respondents by sex in Figure 4. This saves two figures, and should reduce space in the final version.

@@

*PGS effect on fertility = PGS correlation with education X education effect on fertility + direct effect of PGS on fertility*

then indeed PGS effect on fertility, and PGS correlation with education, will be correlated. In that sense, the mediation analysis is just a more precise version of Figure 6 (more precise because it also estimates the education effect on fertility). We're not sure if there is an obvious way to work out "how much" of the Figure 6 correlation is explained by mediation. It may be a bit like asking "how much of a bivariate correlation is explained by the same correlation but with controls": these are really two alternate models of the same process, one less controlled than than the other. We are probably hitting the limits of our econometric expertise, so we are open to suggestions.

> \> In the mediation analysis, why do the authors only use education as the mediator? The mediation framework they use can easily accommodate many mediators (see Section 6 of the Supplementary Information of the "EA2" paper by Okbay et al., Nature Genetics 2016). I would like to see a more full-fledged mediation analysis that also includes other proxies for human capital as mediators, including earnings in a respondent's first job as well as household income (though the later is not an ideal variable for that, especially for females).

We would like to have more mediators, but appropriate variables are hard to find. Earnings in first job reduces our N by a factor of 10, which makes results too noisy to be informative. As the reviewer suggests, current household income is very much endogenous not only to human capital, but also to previous fertility, which makes it inappropriate as a mediator.

@@

> \> Also, in the mediation analysis, some of the ratios of indirect to total effects are negative or larger than 100%. This could indicate that the indirect and direct effects have opposite signs, but could also be due to the imprecision of the estimates. The authors should estimate 95% CIs for the ratio of indirect/total effects (these can be obtained by bootstrapping). They could then focus their discussion on the ratios that are significant and not on those that are just noisy estimates.

Bootstrapping ratios is not so simple, since if the confidence interval for the denominator contains 0, the ratio confidence set may contain infinity and may not be a single interval, and the standard approach of taking quantiles can be misleading (see Franz 2007). Alternatives like the Fieller method have the same problem. Another issue is that to estimate Bonferroni-corrected significance at p = 0.05/33 we'd need 0.00075- and 0.99925-quantiles of the distribution, which would require many thousands of bootstraps (unfeasibly slow, given our sample size).

@@

> 7\. The authors write that sigma captures risk aversion in income in their model. But that's confusing and not a good way of describing this for their model, because as far as I understand there's no income risk! They should clarify this and say that sigma regulates the curvature of the utility function, and explain what this means in their riskless model.

It is a fair point. We have rewritten accordingly: "$\sigma > 0$ measures the curvature of the utility function, i.e. the decline in marginal utility of income as income increases".\
\
OTHER COMMENTS AND SUGGESTIONS

> 8\. Most figures are hard to read. For example, Figures 3-5 are hard to read. In Figure 3a, the dot color indicate education and the dot border color indicates significance; that's very difficult to read quickly (for instance, a blue dot with a yellow border looks green). Instead, the authors could use different symbols (crosses, squares) for education and colors for significance, or vice-versa.

We experimented a lot with our basic figures, and they may not be easy to improve. The graphs' key message is "effect sizes in one category are larger than others". To see that the eye needs to be able to quickly differentiate between categories. We have tried several variants using symbols, but in our judgment the added visual noise makes it harder to follow the categories across rows. The figure below shows three alternative versions: our original, a version using symbols, and one with subtle tweaks to colors and stroke sizes, which we hope fix the greenness issue and make p-values clearer. We prefer our tweaked version (and have used it in the revision), but we will bow to the reviewers or editor.

We've worked on Figures 1 and 2 to improve readability.

```{r different-plots, echo = FALSE, fig.cap = "Alternative plot styles", fig.subcap = c("Original", "Symbols", "Tweaked"), fig.ncol = 1, fig.height=2.5, fig.align="center"}

### 2022-03-15 why-natural-selection.Rmd
Response to BG reviewers:   - tweak colors and graphs   - don't split Fig 3 by sex   - make section 3 briefer   - unify tables of change over time

  fill_colors <- c("steelblue4", "#D63C3C")

@@

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Green triangles show a significant linear increase over time (p < 0.05/33). Red squares show a significant decrease."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        fatten = 1.25) +
      facet_wrap(~score_name, ncol = 5) +
      my_hline +
      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c("-" = "square", "+" = "triangle", "o" = "circle open")) +
      theme(
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

[^normalization]:The selection effect $\beta$ equals $Cov(RLRS, PGS)/Var(PGS)$
where $N$ is the number of children. Since PGS are normalized to variance 1 and
mean 0, this reduces to 
$Cov(RLRS, PGS) = E(RLRS\times PGS) - E(RLRS) E(PGS) = E(RLRS \times PGS)$. This is the polygenic 
score weighted by relative lifetime reproductive success, which is the average polygenic score 
in the next generation [@robertson1966mathematical].

@@

res_wt_combined %>% 
      mutate(
        Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                    "Age/Qual/AFLB", "None"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.5, stroke = 0.5) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "None"              = "grey15", 
        "Geographical"      = "dodgerblue4",
        "Age/Qualification" = "steelblue2",
        "Age/Qual/AFLB"      = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "None" = "square filled", 
        "Geographical" = "circle filled",
        "Age/Qualification" = "circle filled", 
        "Age/Qual/AFLB" = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
      my_theme()

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income, education and sex, and
family structure variables including age at first live birth, presence of a partner, and
lifetime number of sexual partners.

@@

Figure \@ref(fig:plot-income-educ-level) plots selection effects for each
polygenic score, grouping respondents by age of completing full-time education,
and by household income. Effects are larger and more significant for the lowest
income category, and for the lowest education category. These results are robust 
to controlling for respondents' age (Appendix section \@ref(sec:age-control)).

@@

We next focus on variables related to household type and reproductive
strategy. We split males and females by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
For both sexes, selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).[^sex]

[^sex]: We also compared selection effects between men and women (Appendix Figure
\@ref(fig:plot-sexes)).

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
separately for each tercile of AFLB. Several effects are strikingly different
across terciles. ADHD and MDD are selected for amongst the youngest third of
mothers, but selected against among the oldest two-thirds. Educational
attainment is selected for among the oldest two-thirds of mothers, but is not
significantly selected among the youngest third. Similarly, several polygenic
scores for body measurements are selected against only among older mothers. The
correlation between effect sizes for the youngest and oldest terciles is 
`r cor_age_flb`.

@@

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and household type.", fig.subcap = c("Lifetime number of sexual partners", "Household type"), fig.ncol = 1, fig.align = "center", fig.height = 4, fig.width = 4.5}

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(-5, 0, 0, -85),
                  legend.margin        = margin(0, 0, 0, 0),
                  legend.box.just      = "left",
                  legend.justification = "left",
                  legend.spacing       = unit(1, "lines")
                )
sexes_guides <- guides(
                  # fill  = guide_legend(ncol = 1),
                  # color = guide_legend(ncol = 1)
                )

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = `N partners`, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
        sexes_guides

res_with_partner_sex %>% 
      mutate(
        score_name = pretty_names(score_name),
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = Household, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(name = "", values = c("95% c.i." = 1)) +
        sexes_guides

@@

@@

------------------------------------------------------------------------------------
    **Theory: the fertility-human capital           **Empirical results**
    relationship is...**                                      
--- ----------------------------------------------- --------------------------------
1.  Negative: $\frac{dN^{*}}{dh} < 0$.              Figures \@ref(fig:plot-means-over-time) 
                                                    and \@ref(fig:plot-res-wt).

2.  Weaker (closer to zero) at higher wages and/or  Figure \@ref(fig:plot-income-educ-level)a.
    levels of human capital.                        Selection effects are also weaker at 
                                                    higher polygenic scores for 
                                                    educational attainment (Appendix Figure
                                                    \@ref(fig:plot-purifying)).

3.  More negative when the time burden              Stronger effects for single parents
    of children, $b$, is larger.                    (Figure \@ref(fig:plot-n-partners)).

4.  Weaker at higher levels of education ($s$).     Figure \@ref(fig:plot-income-educ-level)b.

----------------------------------------------------------------------------------------

@@

dfr %>% 
      mutate(
        Controls = fct_relevel(Controls, "None"),
        PGS      = ifelse(abs(effect_size) > 0.035, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      filter(
        Controls == "None"
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.005) +
        facet_wrap( ~ Correlation, scales = "free") + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        coord_cartesian(clip = "off") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(
          legend.position = "none",
          panel.spacing = unit(20, "pt")
        )

@@

Second, for the `r nrow(res_mediation)` scores where the bivariate correlation in
\eqref{eq:regression} is significant at $p < 0.05/33$, we run a mediation
analysis to directly test whether the correlation between each polygenic score
and fertility is mediated by educational attainment (Appendix Table
\@ref(tab:tbl-res-mediation)). For `r n_pos_sig` scores, the indirect effect of
the score on fertility *via* educational attainment takes the same sign as the
overall effect, and is significantly different from zero ($p < 0.05/33$). Among
these scores, the median proportion of the total effect explained by the
indirect effect is `r percent(median_prop)`. Figure \@ref(fig:plot-mediation)
shows estimated proportions explained along with bootstrap 95% confidence
intervals (uncorrected; 100 bootstraps). The educational attainment variable is
a relatively crude measure of human capital: more accurate measures would likely
explain more of the total effect.

@@

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/33` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/33`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i.",
                        ),
                        width = 0.3,
                        color = "grey30"
                      ) +
        scale_fill_manual(values = pval_palette) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))

@@

## Natural selection by sex

@@

Stabilizing selection reduces variance in the trait under selection, while
disruptive selection increases variance. To check for these, we rerun equation
\@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
age/qualification weights. Scores for hip circumference show significant
stabilizing selection ($p < 0.05/33$, negative coefficient on quadratic term).
The EA2 score for educational attainment shows significant disruptive
selection ($p < 0.05/33$, positive coefficient), which reduces the strength of
selection against educational attainment at very high levels of the PGS. Figure
\@ref(fig:plot-purifying) plots predicted number of children against polygenic
score from these regressions.

@@

\FloatBarrier
\clearpage
## Parents' generation {#sec-parents}
### Selection effects

@@

Tables \@ref(tab:tbl-siblings-by-period) and \@ref(tab:tbl-children-by-period)
summarize the results. We report the number of scores showing significant
changes over time (i.e. a significant interaction between polygenic score and
the "late born" dummy): either a significant change in sign, a significant
increase in size, or a significant decrease in size. There is little evidence
for changes in selection effects within the parents' generation, with just one
score showing a significant decrease in size. In the respondents' generation,
effect sizes were significantly larger in absolute size among the later-born for
nine polygenic scores: ADHD, age at menopause, caffeine consumption, cognitive
ability, coronary artery disease, EA2, EA3, extraversion and major depressive
disorder. These changes are inconsistent with the intergenerational change,
where estimated effect sizes were larger among the earlier, parents' generation.
One score, for alzheimer's disease, showed a significant change in sign, from
negative to positive effects on fertility.

@@

```{r tbl-siblings-by-period}

period_table <- function (tbl, caption) {
  tbl %>% 
      as_hux() %>%
      set_bold(1, everywhere) %>% 
      set_bottom_border(1, everywhere) %>%

      add_footnote(glue("Significance is measured at p < 0.05/{n_tests}"), 
        border = 0.4, font_size = 8) %>% 
      set_contents(1, everywhere, 
        c("Change", "Number of scores")) %>% 
      set_right_border(final(1), 1, 0) %>% 
      set_top_padding(1) %>% 
      set_bottom_padding(1) %>%
      set_left_padding(everywhere, 1, 0) %>% 
      set_right_padding(everywhere, 2, 0) %>% 
      set_caption(caption) %>% 
      set_caption_width(1)
}

>>>>

  fill_colors <- c("steelblue4", "darkred")

@@

```{r plot-means-over-time, fig.height = 6, fig.cap = "Mean polygenic scores (PGS) by birth year in UK Biobank. Points are means for 5-year intervals. Lines are 95% confidence intervals. Circles show a significant linear increase (green) or decrease (red) over time (p < 0.05/33)."}

@@

pgs_over_time %>% 
      mutate(score_name = pretty_names(score_name)) %>% 
      ggplot(aes(YOB, group = score_name, colour = Change, shape = Change)) + 
      geom_pointrange(aes(ymin = score_lo, y = score, ymax = score_hi), 
        fatten = 1.2) +
      facet_wrap(~score_name, ncol = 6) +
      my_hline +
      scale_color_manual(values = c("-" = "red", "+" = "darkgreen", "o" = "black")) +
      scale_shape_manual(values = c("-" = "circle", "+" = "circle", "o" = "square")) +
      theme(
        line        = element_line(size = 0.5, lineend = "round"),
        strip.text  = element_text(size = 6), 
        axis.text.x = element_text(size = 7),
        legend.position = "none"
      ) +
      labs(y = "PGS", x = "Birth year (5 year intervals)")

@@

[^normalization]:The selection effect $\beta$ equals $Cov(RLRS, PGS)/Var(PGS)$.
Since PGS are normalized to variance 1 and mean 0, this reduces to $Cov(RLRS,
PGS) = E(RLRS\times PGS) - E(RLRS) E(PGS) = E(RLRS \times PGS)$. This is the
polygenic score weighted by relative lifetime reproductive success, which is the
average polygenic score in the next generation [@robertson1966mathematical].

@@

res_wt_combined %>% 
      mutate(
        Weights  = fct_relevel(Weights, "Geographical", "Age/Qualification",
                    "Age/Qual/AFLB", "None"),
        term     = pretty_names(term),
        term     = fct_reorder(term, estimate, order_abs(1))
      ) %>% 
      rename(p = p.value) %>% 
      # can't use standard_ggplot because we have non-standard geom_point()
      ggplot(aes(
        x      = estimate, 
        y      = term, 
        yend   = term,
        shape  = Weights, 
        fill   = Weights, 
        colour = p < 0.05/{{n_regs}}
      )) +
      my_vline +
      geom_point(size = 1.6, stroke = 0.8, alpha = 0.9) +
      confint_geom_segment() +
      pval_scale() +
      scale_fill_manual(values = c(
        "None"              = "grey15", 
        "Geographical"      = "dodgerblue4",
        "Age/Qualification" = "steelblue2",
        "Age/Qual/AFLB"      = "darkseagreen4"
      )) +
      scale_shape_manual(values = c(
        "None" = "square filled", 
        "Geographical" = "circle filled",
        "Age/Qualification" = "circle filled", 
        "Age/Qual/AFLB" = "circle filled"
      )) + 
      labs(x = "Effect size", y = "") +
      scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
      my_theme()

@@

To motivate our economic model, we split the sample by demographic
and social variables, and re-estimate \@ref(eq:regression) within each subgroup.
We examine socio-demographic variables including income and education, and
family structure variables including age at first live birth, presence of a
partner, and lifetime number of sexual partners. Figure
\@ref(fig:plot-income-educ-level) plots selection effects for each polygenic
score, grouping respondents by age of completing full-time education, and by
household income. Effects are larger and more significant for the lowest income
category, and for the lowest education category. These results are robust to
controlling for respondents' age (Appendix section \@ref(sec:age-control)).

@@

Turning to family structure, we split respondents by lifetime number of sexual
partners, at the median value of 3 (Figure \@ref(fig:plot-n-partners)a).
Selection effects are larger and more significant among
those with more than 3 lifetime partners. Next we split respondents by whether
they were living with a spouse or partner at the time of interview.
Effects are larger among those not living with a spouse or partner
(Figure \@ref(fig:plot-n-partners)b).[^sex]

[^sex]: The same pattern holds if we analyse men and women separately (Appendix Figure
\@ref(fig:plot-n-partners-sex)). We also directly compared selection effects 
between men and women (Appendix Figure \@ref(fig:plot-sexes)).

```{r plot-n-partners, fig.cap = "Selection effects by number of sexual partners and presence of a partner.", fig.subcap = c("Lifetime number of sexual partners", "Presence of a partner"), fig.ncol = 1, fig.align = "center", fig.height = 4}

drake::loadd(res_partners_joint)
res_partners_joint %<>%       
      filter(grepl(":", term)) %>% 
      mutate(`N partners` = ifelse(grepl("TRUE:", term), "3 or less", "4 or more"))

n_regs <- as.double(nrow(res_partners_joint))
res_partners_joint %>% 
      standard_ggplot(fill_col = `N partners`, n_regs = n_regs, 
                        score_col = score_name)

drake::loadd(res_with_partner)
res_with_partner %<>% filter(grepl(":", term))

n_regs <- as.double(nrow(res_with_partner))

res_with_partner %>% 
      mutate(
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      standard_ggplot(fill_col = Household, n_regs = n_regs, 
                        score_col = score_name)
```

@@

Lastly, we split female respondents by age at first live birth
(AFLB).[^aflb-no-men] There is evidence for genetic effects on AFLB
[@Barban_2016], and there is a close link between this variable and number of
children born. Figure \@ref(fig:plot-age-flb) shows effect sizes estimated
separately for each tercile of AFLB. Effects are strikingly different
across terciles. Educational attainment, ADHD and MDD are selected for amongst
the youngest third of mothers, but selected against among the oldest two-thirds.
Similarly, several polygenic scores for body measurements are selected against
only among older mothers. The correlation between effect sizes for the youngest
and oldest terciles is `r cor_age_flb`.

@@

@@

------------------------------------------------------------------------------------
    **Theory: the fertility-human capital         **Empirical results**
    relationship is...**                                      
--- --------------------------------------------- ----------------------------------
1.  Negative: $\frac{dN^{*}}{dh} < 0$.            Figures \@ref(fig:plot-means-over-time) 
                                                  and \@ref(fig:plot-res-wt).

2.  Weaker (closer to zero) at higher wages       Figure 
    and/or levels of human capital.               \@ref(fig:plot-income-educ-level)a.
                                                  Selection effects are also weaker at
                                                  higher polygenic scores for 
                                                  educational attainment 
                                                  (Appendix Figure
                                                  \@ref(fig:plot-purifying)).

3.  More negative when the time burden            Stronger effects for single parents
    of children $b$ is larger.                    (Figure \@ref(fig:plot-n-partners)).

4.  Weaker at higher levels of education $s$.     Figure \@ref(fig:plot-income-educ-level)b.

------------------------------------------------------------------------------------

@@

@@

dfr %>% 
      mutate(
        PGS      = ifelse(abs(effect_size) > 0.02, PGS, ""),
        PGS      = sub("(^\\w*?)_.*", "\\1", PGS)
      ) %>% 
      filter(
        Controls == "None"
      ) %>% 
      ggplot(aes(effect_size, cor, colour = Correlation)) + 
        geom_point() +
        geom_text(aes(label = PGS), size = 3, colour = "black", 
                    check_overlap = TRUE, nudge_x = 0.002) +
        facet_wrap( ~ Correlation, scales = "free") + 
        my_vline + 
        my_hline +
        labs(x = "Selection effect", y = "Correlation") +
        coord_cartesian(clip = "off") +
        scale_colour_brewer(type = "qual", palette = 2) +
        theme(
          legend.position = "none",
          panel.spacing = unit(20, "pt")
        )

@@

Second, for the `r nrow(res_mediation)` scores where the bivariate correlation in
\eqref{eq:regression} is significant at $p < 0.05/33$, we run a mediation
analysis to directly test whether the correlation between each polygenic score
and fertility is mediated by educational attainment (Appendix Table
\@ref(tab:tbl-res-mediation)). Figure \@ref(fig:plot-mediation) shows estimated
proportions explained along with bootstrap 95% confidence intervals
(uncorrected; 100 bootstraps). For `r n_pos_sig` scores, the indirect effect of
the score on fertility *via* educational attainment takes the same sign as the
overall effect, and is significantly different from zero ($p < 0.05/33$). Among
these scores, the median proportion of the total effect explained by the
indirect effect is `r percent(median_prop)`. The educational attainment variable
is a relatively crude measure of human capital: more accurate measures would
likely explain more of the total effect.

@@

res_mediation %>% 
      mutate(
        term = pretty_names(term),
        term = fct_reorder(term, prop_ind, order_abs(1)),
        `Indirect effect p < 0.05/33` = sig 
      ) %>% 
      ggplot(aes(y = term)) + 
        geom_col(aes(x = prop_ind, fill = `Indirect effect p < 0.05/33`), 
                   width = .5) + 
        geom_errorbar(aes(
                          xmin = prop_ind_conf_low, 
                          xmax = prop_ind_conf_high,
                          linetype = "Bootstrap 95% c.i.",
                        ),
                        width = 0.3,
                        color = "grey50"
                      ) +
        scale_fill_manual(values = pval_palette) +
        my_vline +
        scale_x_continuous(labels = scales::percent, n.breaks = 7) +
        coord_cartesian(xlim = c(-0.5, 1)) +
        labs(x = "Proportion mediated", y = "") +
        scale_linetype_manual(name = "", values = c("Bootstrap 95% c.i." = 1))

@@

## Selection effects by sex

@@

Stabilizing selection reduces variance in the trait under selection, while
disruptive selection increases variance. To check for these, we rerun equation
\@ref(eq:regression), adding a quadratic term in $PGS_i$, and using our
age/qualification weights. Scores for hip circumference show significant
stabilizing selection ($p < 0.05/33$, negative coefficient on quadratic term).
The EA2 score for educational attainment shows significant disruptive selection
($p < 0.05/33$, positive coefficient), which reduces the strength of selection
against educational attainment at very high levels of the PGS. (The quadratic on
the EA3 score has a similar coefficient but is not significant at $p < 0.05/33$.)
Figure \@ref(fig:plot-purifying) plots predicted number of children
against polygenic score from these regressions.

@@

\FloatBarrier
\clearpage
## Number of partners and presence of partner by sex

Figure \@ref(fig:plot-n-partners-sex) splits up Figure \@ref(fig:plot-n-partners)
by sex. The pattern of results is the same in both sexes: selection effects are stronger among those with more lifetime sexual partners, and among those not
currently living with a partner.

```{r plot-n-partners-sex, fig.cap = "Selection effects by number of sexual partners and presence of a partner, for men and women separately.", fig.subcap = c("Lifetime number of sexual partners", "Presence of a partner"), fig.ncol = 1, fig.align = "center", fig.height = 4}

sexes_theme <-  my_theme(
                  plot.margin          = unit(c(1, 1, 0, 1), "lines"),
                  legend.position      = "bottom", 
                  #legend.box.spacing   = unit(0, "lines"),
                  legend.box.margin    = margin(-5, 0, 0, -85),
                  legend.margin        = margin(0, 0, 0, 0),
                  legend.box.just      = "left",
                  legend.justification = "left",
                  legend.spacing.y     = unit(0, "lines"),
                  #legend.spacing       = unit(0.5, "lines"),
                  legend.title         = element_text(size = 10)
                  #legend.text          = element_text(size = 8)
                )
sexes_guides <- guides(
                  fill  = guide_legend(ncol = 1),
                  color = guide_legend(ncol = 1)
                )

n_regs <- as.double(nrow(res_partners))
res_partners %>% 
      mutate(
        score_name = pretty_names(score_name)
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = `N partners`, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(name = "", values = c("95% c.i. uncorrected" = 1)) +
        sexes_guides

res_with_partner_sex %>% 
      mutate(
        score_name = pretty_names(score_name),
        Household = ifelse(grepl("TRUE", term), "With partner", 
                      "Without partner")
      ) %>% 
      rename(p = p.value) %>% 
      ggplot(aes(
        y      = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        yend   = reorder_within(score_name, estimate, sex, fun = order_abs(1)), 
        x      = estimate, 
        fill   = Household, 
        colour = p < 0.05/{{n_regs}}
      )) + 
        my_vline +
        confint_geom_segment() +
        pval_geom_point() +
        facet_wrap(vars(sex), scales = "free_y") +
        my_fill_scale(n = 2) +
        pval_scale() +
        sexes_theme +
        my_labs(x = "", y = "") + 
        scale_y_reordered() +
        scale_linetype_manual(
                                name   = "", 
                                values = c("95% c.i. uncorrected" = 1)
                              ) +
        sexes_guides

\FloatBarrier
\clearpage
## Parents' generation {#sec-parents}
### Selection effects and change over time

@@

Table \@ref(tab:tbl-change-by-period) 
summarizes the results. We report the number of scores showing significant
changes over time (i.e. a significant interaction between polygenic score and
the "late born" dummy): either a significant change in sign, a significant
increase in size, or a significant decrease in size. There is little evidence
for changes in selection effects within the parents' generation, with just one
score showing a significant decrease in size. In the respondents' generation,
effect sizes were significantly larger in absolute size among the later-born for
nine polygenic scores: ADHD, age at menopause, caffeine consumption, cognitive
ability, Coronary Artery Disease, EA2, EA3, extraversion and Major Depressive
Disorder. These changes are inconsistent with the intergenerational change,
where estimated effect sizes were larger among the earlier, parents' generation.
One score, for alzheimer's disease, showed a significant change in sign, from
negative to positive effects on fertility.

@@

```{r tbl-change-by-period}

tbl_resp <- summary_period %>% 
              group_by(Change_Respondents) %>% 
              count()

tbl_par <- summary_period %>% 
             group_by(Change_Parents) %>% 
             count()

tbl_both <- full_join(tbl_par, tbl_resp, 
                        by = c("Change_Parents" = "Change_Respondents"))
